---
title: "Harvard CYO Boston housing"
author: "Rekha Rao"
date: "4/11/2023"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}

```
1. Introduction

We will be using the Boston housing dataset which is one of the most popular datasets to be used for machine learning.

1.a. Structure of the Report


Section 1 - Introduction and Objective


Section 2 - Data Preparation


Section 3 - Exploratory Data Analysis 


Section 4- Development of the models


Section 5 - Conclusion


Section 6 - Limitations


Section 7 - References



1.b. Objective 


The objective is to build a machine learning model to predict the Median value (MEDV) of owner occupied homes in Boston based on the available features. 



1.c. Overview


The data set has been downloaded from Kaggle. 

Since Kaggle does not allow us to download the files directly, I have downloaded the file to my github and here is the link to the file:

https://github.com/rrao2511/CYO-Harvard-Capstone-Project/raw/main/housing.csv


The original Boston housing dataset contains 506 samples and 14 variables. 


For the purpose of this report we will be looking only at a subset of the original Boston housing dataset. 


Our dataset contains  489 samples and 4 variables which are explained below:

MEDV – 	Median Value of Owner occupied homes

RM   -      	Average number of rooms per dwelling

LSTAT  -   	% lower status of population

PT RATIO - 	Pupil teacher ratio by town

The goal of our analysis is to select the best prediction model which can predict the Median value of owner occupied homes in Boston.



1.d. Approach 

The steps we will take for this project are:


Look at the data structure


Data Preparation and Cleaning


Exploring the data 


Development of models


Results 


Recommendation of the Model


Limitations of the Model




Start of Script 


2. Data Preparation and Data Preprocessing


 Dataset Source:

 Kaggle is an online platform for data scientists and machine learning students.

 This particular dataset- Boston housing has been downloaded from Kaggle. 


 First step - download the Packages needed for this analysis and load the libraries.


```{r}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(ggplot2)) install.packages("ggplot2", repos = "http://cran.us.r-project.org")
if(!require(dplyr)) install.packages("dplyr", repos = "http://cran.us.r-project.org")
if(!require(corrplot)) install.packages("corrplot", repos = "http://cran.us.r-project.org")
if(!require(randomForest)) install.packages("randomForest", repos = "http://cran.us.r-project.org")
if(!require(rpart)) install.packages("rpart", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(rpart.plot)) install.packages("rpart.plot", repos = "http://cran.us.r-project.org")
if(!require(e1071)) install.packages("e1071", repos = "http://cran.us.r-project.org")


library(tidyverse)
library(ggplot2)
library(caret)
library(dplyr)
library(corrplot)
library(randomForest)
library(rpart)
library(rpart.plot)
library(e1071)

```


Download the dataset 

Since Kaggle does not allow us to download the files directly, have downloaded the file to my github and here is the link to the file:

https://github.com/rrao2511/CYO-Harvard-Capstone-Project/raw/main/housing.csv


Reading the data from the csv file


```{r}
boston_housing<-read.csv("https://github.com/rrao2511/CYO-Harvard-Capstone-Project/raw/main/housing.csv",header=TRUE,sep=",",quote ="\"")

```


For the purpose of this analysis we are looking at a subset of the Boston housing set

 First lets look at the data set - checking the dimension.
 
 
 This dataset has 489 observations and 4 columns. This is a
 subset of the original Kaggle dataset.
 
 There are 4 columns and the details of the column are shown
 below.We will be using all the 4 columns for our analysis.
 
 
 We will also look at the structure of the dataset, head and
 the summary.
 
 
 Explanation of Column names and details 

 RM - Average number of rooms per dwelling

 LSTAT - % lower status of population

 PT Ratio - Pupil teacher ratio by town

 MEDV - Median Value of owner occupied homes in $1000s.



```{r}
dim(boston_housing)

str(boston_housing)

head(boston_housing)

summary(boston_housing)

```


Cleaning up the data 

 Since this dataset is already clean, data cleaning was not needed and it could be used directly for analysis.


 Check to see if there are duplicate values and also any missing values.
```{r}
sum(duplicated(boston_housing))


sum(is.na(boston_housing))
```


3. Exploratory Data Analysis using Data Visualization

Before we start building the model we will understand the data set by doing some Exploratory Data Analysis.


 Check the correlation between variables by plotting a correlation graph


```{r}
corrplot(cor(boston_housing), method = "number", type = "upper", diag = FALSE)
```
From correlation matrix, we observe that:


 Both RM and LSTAT have a strong correlation with MEDV. 

 Median value of owner-occupied homes (in 1000$) increases as average number of rooms per dwelling increases and it decreases if percent of lower status population in the area increases.

 PT Ratio has a positive correlation with LSTAT


 Next lets look at Scatter plots to show relationship between
 Median value and variables



```{r}
boston_housing%>%
  gather(key, val,-MEDV) %>%
  ggplot(aes(x = val, y = MEDV/1000))+
  geom_point()+
  stat_smooth(method = "lm", se = TRUE, col ="blue") +
  facet_wrap(~key, scales = "free")+
  theme_grey()+
  ggtitle("Scatter plot - Dependent variables vs Median value(medv)")
```


From the plots we see that RM and LSTAT have a strong correlation with Median value.

The Median value prices increases as the RM value increases linearly.

The Median value prices tend to decrease with an increase in LSTAT



4. Developing the Models

 We will use three different models for this project:

 Decision trees, Random Forest and Support Vector Machine.

 We will evaluate the models using Root Mean Squared Error (RMSE).

 The model that best fits the data will be selected.

 First we need to split the data into train sets and test
 sets:

 Data is split into train and test sets - 80:20 



```{r}
set.seed(123)
bh_index<- sample(nrow(boston_housing),nrow(boston_housing)*.80)
bh_train<- boston_housing[bh_index,]
bh_test<- boston_housing[-bh_index,]
```


 Model Development

 Model 1 - Decision trees


 Steps - we will create the model using Decision trees on the

 train set, plot  the decision tree, validate on the test set
 
 and finally calculate the RMSE. 


```{r}
bhtree.fit<- rpart(MEDV~., data= bh_train)

rpart.plot(bhtree.fit, type = 3, box.palette = c("red", "green"), fallen.leaves = TRUE)

tree.pred<- predict(bhtree.fit, newdata = bh_test)

tree.rmse<- sqrt(mean((bh_test$MEDV- tree.pred)^2))

cat("Decision Tree RMSE", round(tree.rmse,2),"\n")
```


Model 2 -  Random Forest

 Steps - we will create the model using Random forest on train
 
 set, validate on the test set and finally  calculate the

 RMSE.


```{r}
rf.fit<- randomForest(MEDV~., data= bh_train, ntree= 500, mtry = 3)

rf.pred<- predict(rf.fit, newdata = bh_test)

rf.rmse<- sqrt(mean((bh_test$MEDV - rf.pred)^2))


cat("Random Forest RMSE", round(rf.rmse,2), "\n")
```

Model 3 -  Support Vector Machines (SVM)

 Steps - we will create the model using SVM on train
 
 set, validate on the test set and finally  calculate the

 RMSE.


```{r}
svm.fit<- svm(MEDV~., data = bh_train, kernel= "linear", cost =1)

svm.pred<- predict(svm.fit, newdata = bh_test)

svm.rmse<- sqrt(mean((svm.pred- bh_test$MEDV)^2))


cat("SVM RMSE:", svm.rmse, "\n")

```


5. Conclusion


Create a table for the RMSE values of Decision trees, Random Forest and SVM


```{r}
results_table<- data.frame(Model = c("Decision Tree", "Random Forest", "SVM"),
                         RMSE= c(tree.rmse,rf.rmse,svm.rmse ))
```


 Based on the above results here are our observations:

 a) The random forest model has the lowest RMSE value of  
    71805,indicating that it may be the best model for
    predicting the median value of owner-occupied homes in
    Boston.

 b) Whereas the Decision tree and the SVM models have higher
    RMSE values of 79926 and 93156 respectively, indicating
    that they may not be the best choice for predicting the
    median value of owner occupied homes.




 6. Limitations of the Model

 We need to be cautious we need to be cautious when drawing 
 conclusions based on RMSE values alone, as there may be other
 factors to consider such as model complexity, 
 interpretability, and computational efficiency

 Random forest models can be further improved with
 hyperparameters tuning.
 
 But on account of shortage of time this was not attempted. 

 Similarly the SVM model could be tuned further by changing
 the parameters and the kernel function.
 
 But on account of shortage of time this was not attempted. 


=========== END OF FILE ====================================
